---
title: "Class 9 practice"
author: "CouBC"
date: '2018-10-02'
output: 
  html_document: 
    keep_md: yes
---

```{r}
library(tidyverse)
```


dat <- data.frame(
   id = LETTERS[1:5],
   x = sample(0:1, 5, TRUE),
   y = sample(0:1, 5, TRUE),
   z = sample(0:1, 5, TRUE)
)
```{r}
dat <- data.frame(id = LETTERS[1:5])
print(dat)
```

```{r}
#sample generates random letters - need to add TRUE for it to work

dat <- data.frame(id = LETTERS[1:5], x = sample(0:1, 5, TRUE))
print(dat)
```

```{r}
dat <- data.frame(id = LETTERS[1:5], x = sample(0:1, 5, TRUE), y = sample(0:1, 5, TRUE), z = sample(0:1, 5, TRUE))
print(dat)
```


rnorm() generates random numbers pos/and neg but needs three items in its argument which I dont understand
e.g., rnorm(1, 2, 3)


Gather in action
dat %>% gather(item, scores, -c(id))

```{r}
dat %>% 
  gather(item, scores, -c(id)) %>% 
  knitr::kable()
```

Spread it out:

```{r}
dat %>% gather(item, scores, -c(id)) %>%
    spread(item, scores)
```

```{r}
#alternatively:

dat %>% gather(item, scores, x:z) %>% 
knitr::kable()
```

```{r}
#or this works too:
dat %>% gather(item, scores, x, y, z) 
```

dat %>% gather(item, scores, -c(id)) %>%
    spread(id, scores)
    
or 

dat %>% gather(item, scores, -c(id)) %>%
    spread(item, scores)
    
  Read these three LOTR csv's, saving them to `lotr1`, `lotr2`, and `lotr3`:

- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv
- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv
- https://github.com/jennybc/lotr-tidy/blob/master/data/The_Return_Of_The_King.csv

```{r}
lotr1 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv")
lotr2 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv")
lotr3 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv")
```

```{r}
lotr1
```

```{r}
lotr2

```

```{r}
lotr3
```

```{r}
lotr.untidy <- bind_rows(lotr1, lotr2, lotr3) 
```


```{r}
str(lotr.untidy)
```

```{r}
lotr.untidy
```



```{r}
#Jenny's way
lotr.tidy <- gather(lotr.untidy, key = "Gender", value = 'Words', Female, Male )
lotr.tidy

```



```{r}
lotr.tidy %>% 
  count(Gender, Race, wt = words)
```

```{r}
table1
#> # A tibble: 6 x 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```

```{r}
table2 
#> # A tibble: 12 x 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
```

```{r}
table3
#> # A tibble: 6 x 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583

```

```{r}
table4a  # cases
#> # A tibble: 3 x 3
#>   country     `1999` `2000`
#> * <chr>        <int>  <int>
#> 1 Afghanistan    745   2666
#> 2 Brazil       37737  80488
#> 3 China       212258 213766

table4b  # population
#> # A tibble: 3 x 3
#>   country         `1999`     `2000`
#> * <chr>            <int>      <int>
#> 1 Afghanistan   19987071   20595360
#> 2 Brazil       172006362  174504898
#> 3 China       1272915272 1280428583
```

```{r}
table4a
```

```{r}
table4b
```

```{r}
table2
```



Compute the rate for table2, and table4a + table4b. You will need to perform four operations:

    Extract the number of TB cases per country per year.
    Extract the matching population per country per year.
    Divide cases by population, and multiply by 10000.
    Store back in the appropriate place.

Which representation is easiest to work with? Which is hardest? Why?

For Table 2, we need to first create separate tables for cases and population and ensure that they are sorted in the same order.

t2_cases <- filter(table2, type == "cases") %>%
  rename(cases = count) %>%
  arrange(country, year)
t2_population <- filter(table2, type == "population") %>%
  rename(population = count) %>%
  arrange(country, year)

```{r}
#Extract the number of TB cases per country per year.

t2_cases <- filter(table2, type == "cases") %>% 
  rename(cases = count) %>% 
  arrange(country, year)

# Extract the matching population per country per year.

t2_population <- filter(table2, type == "population") %>% 
  rename(population = count) %>%
  arrange(country, year)

t2_population
t2_cases

```

Calculate the cases per capita in a separate data frame.

t2_cases_per_cap <- t2_cases %>%
  mutate(population = t2_population$population,
         cases_per_cap = (cases / population) * 10000) %>%
  select(country, year, cases_per_cap)

```{r}
#Divide cases by population, and multiply by 10000.

t2_cases_per_cap <- t2_cases %>% 
  mutate(population = t2_population$population,
         cases_per_cap = (cases / population) * 10000) %>% 
  select(country, year, cases_per_cap)

t2_cases_per_cap

```

```{r}
#For table4a and table4b, we will create a separate table for cases per capita (table4c), with country rows and year columns.

tibble(country = table4a$country,
       `1999` = table4a[["1999"]] / table4b[["1999"]] * 10000,
       `2000` = table4a[["2000"]] / table4b[["2000"]] * 10000)

```

```{r}

#Table2 is most complex to calculate, Table4a Table4b better, but if we tried this from Table1 we could caclulate with mutate

table1 %>% 
  mutate(cases_per_capita = cases / population * 10000)
```

```{r}
#gathering table4a

table4a %>% 
  gather("1999", "2000", key = year, value = count)
```

To combine the tidied versions of table4a and table4b into a single tibble, we need to use dplyr::left_join(), which you’ll learn about in relational data.

tidy4a <- table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
tidy4b <- table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")
left_join(tidy4a, tidy4b)

```{r}
#first tidy each table before joining in a single tibble

tidy4a <- table4a %>% 
  gather("1999", "2000", key = year, value = cases)

tidy4b <- table4b %>% 
  gather("1999", "2000", key = year, value = population)
```

```{r}
#now join the two tables

left_join(tidy4a, tidy4b)
```

Spreading is the opposite of gathering. You use it when an observation is scattered across multiple rows. For example, take table2: an observation is a country in a year, but each observation is spread across two rows.

```{r}
table2 %>% 
spread(key = type, value = count)
```


```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks

```

```{r}
stocks %>% 
  spread(year, return)
```

```{r}
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)

#notice that year became a character rather than integer after spread - need to use convert = TRUE to retain what it was before. However, since convert = TRUE is guessing the appropriate type it still may not work --> FYI!
```

######NB###
The functions spread() and gather() are not perfectly symmetrical because column type information is not transferred between them. When we use gather() on a data frame, it throws away all the information about the original column types. Additionally, it has to coerce all the variables that are being gathered into a single type, since they are going into a single vector. Later if we spread() that data frame, the spread() function has no way to know what the original data types of the columns that were earlier gathered.

In this example, in the original table, the column year was numeric. After running spread() and gather() it is a character vector. Variable names are always converted to a character vector by gather().

The functions spread() and gather() can be closer to symmetrical if we use the convert argument. It will try to convert character vectors to the appropriate type using type.convert().

```{r}
stocks %>%
  spread(key = "year", value = "return") %>%
  gather(`2015`:`2016`, key = "year", value = "return", convert = TRUE)
```

```{r}
table4a %>%
  gather(`1999`, `2000`, key = "year", value = "cases")

#if no back ticks (` `) around numbers, it will not work, as it is looking for 1999th case and 2000th case
```

```{r}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
#This is a problematic tibble because Phillip woods has 2 ages, 50 and 45 and they both can't occupy the same space - and there is no corresponding height for that age. unclear which height is which age

#therefore: spread(people, key, value) would not work - results in error
```


```{r}
#solution create new column with number of observations

people <- tribble(
  ~name,             ~key,    ~value, ~obs,
  #-----------------|--------|------|------
  "Phillip Woods",   "age",       45, 1,
  "Phillip Woods",   "height",   186, 1,
  "Phillip Woods",   "age",       50, 2,
  "Jessica Cordero", "age",       37, 1,
  "Jessica Cordero", "height",   156, 1
)


```


```{r}
spread(people, key, value)
```

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
```

```{r}
preg_tidy <- preg %>% 
  gather(male, female, key = "sex", value = "count", na.rm = TRUE)

preg_tidy
```

```{r}
#storing pregnant and sex as logical vectors

preg_tidy2 <- preg_tidy %>%
  mutate(female = sex == "female",
         pregnant = pregnant == "yes") %>%
  select(female, pregnant, count)
preg_tidy2

#In the previous data frame, I named the logical variable representing the sex female, not sex. This makes the meaning of the variable self-documenting. If the variable were named sex with values TRUE and FALSE, without reading the documentation, we wouldn’t know whether TRUE means male or female.

#Apart from some minor memory savings, representing these variables as logical vectors results in more clear and concise code.
```

```{r}
#separate() pulls apart one column into multiple columns, by splitting wherever a separator character appears.


```

