---
title: "Class 9 practice"
author: "CouBC"
date: '2018-10-02'
output: 
  html_document: 
    keep_md: yes
---

```{r}
library(tidyverse)
```


dat <- data.frame(
   id = LETTERS[1:5],
   x = sample(0:1, 5, TRUE),
   y = sample(0:1, 5, TRUE),
   z = sample(0:1, 5, TRUE)
)
```{r}
dat <- data.frame(id = LETTERS[1:5])
print(dat)
```

```{r}
#sample generates random letters - need to add TRUE for it to work

dat <- data.frame(id = LETTERS[1:5], x = sample(0:1, 5, TRUE))
print(dat)
```

```{r}
dat <- data.frame(id = LETTERS[1:5], x = sample(0:1, 5, TRUE), y = sample(0:1, 5, TRUE), z = sample(0:1, 5, TRUE))
print(dat)
```


rnorm() generates random numbers pos/and neg but needs three items in its argument which I dont understand
e.g., rnorm(1, 2, 3)


Gather in action
dat %>% gather(item, scores, -c(id))

```{r}
dat %>% 
  gather(item, scores, -c(id)) %>% 
  knitr::kable()
```

Spread it out:

```{r}
dat %>% gather(item, scores, -c(id)) %>%
    spread(item, scores)
```

```{r}
#alternatively:

dat %>% gather(item, scores, x:z) %>% 
knitr::kable()
```

```{r}
#or this works too:
dat %>% gather(item, scores, x, y, z) 
```

dat %>% gather(item, scores, -c(id)) %>%
    spread(id, scores)
    
or 

dat %>% gather(item, scores, -c(id)) %>%
    spread(item, scores)
    
  Read these three LOTR csv's, saving them to `lotr1`, `lotr2`, and `lotr3`:

- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv
- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv
- https://github.com/jennybc/lotr-tidy/blob/master/data/The_Return_Of_The_King.csv

```{r}
lotr1 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv")
lotr2 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv")
lotr3 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv")
```

```{r}
lotr1
```

```{r}
lotr2

```

```{r}
lotr3
```

```{r}
lotr.untidy <- bind_rows(lotr1, lotr2, lotr3) 
```


```{r}
str(lotr.untidy)
```

```{r}
lotr.untidy
```



```{r}
#Jenny's way
lotr.tidy <- gather(lotr.untidy, key = "Gender", value = 'Words', Female, Male )
lotr.tidy

```



```{r}
lotr.tidy %>% 
  count(Gender, Race, wt = words)
```

```{r}
table1
#> # A tibble: 6 x 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```

```{r}
table2 
#> # A tibble: 12 x 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
```

```{r}
table3
#> # A tibble: 6 x 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583

```

```{r}
table4a  # cases
#> # A tibble: 3 x 3
#>   country     `1999` `2000`
#> * <chr>        <int>  <int>
#> 1 Afghanistan    745   2666
#> 2 Brazil       37737  80488
#> 3 China       212258 213766

table4b  # population
#> # A tibble: 3 x 3
#>   country         `1999`     `2000`
#> * <chr>            <int>      <int>
#> 1 Afghanistan   19987071   20595360
#> 2 Brazil       172006362  174504898
#> 3 China       1272915272 1280428583
```

```{r}
table4a
```

```{r}
table4b
```

```{r}
table2
```



Compute the rate for table2, and table4a + table4b. You will need to perform four operations:

    Extract the number of TB cases per country per year.
    Extract the matching population per country per year.
    Divide cases by population, and multiply by 10000.
    Store back in the appropriate place.

Which representation is easiest to work with? Which is hardest? Why?

For Table 2, we need to first create separate tables for cases and population and ensure that they are sorted in the same order.

t2_cases <- filter(table2, type == "cases") %>%
  rename(cases = count) %>%
  arrange(country, year)
t2_population <- filter(table2, type == "population") %>%
  rename(population = count) %>%
  arrange(country, year)

```{r}
t2_cases <- filter(table2, type == "cases") %>% 
  rename(cases = count) %>% 
  arrange(country, year)

t2_cases
```

